---
title: "Code Highlighting Formatting for Verbatim"
format:
  pdf:
    include-in-header: 
      - text: |
          %%%%%%%%%%%%%%%%%%%% USAGE NOTE %%%%%%%%%%%%%%%%%%%%
          % This LaTeX script sets up consistent formatting for DSDE Workshop handouts.
          %
          % To use:
          % 1. Reference this LaTeX script in the YAML header of Quarto documents
          %    with correct relative paths based on their directory depth.
          % 2. Ensure that the "rootdir" and "stylesdir" paths in this LaTeX script
          %    match those relative paths. Adjust as necessary.
          % 
          % The LaTeX script is located in "Code/Handout Formatting/PDF Formatting Script.tex".
          %
          % Example:
          % For:
          %   \input{../Code/Handout Formatting/PDF Formatting Script.tex}
          % Set:
          %   \newcommand{\rootdir}{../Code/Handout Formatting}
          %   \newcommand{\stylesdir}{../_styles/Yale_typeface}
          %
          % For:
          %   \input{../../../../Code/Handout Formatting/PDF Formatting Script.tex}
          % Set:
          %   \newcommand{\rootdir}{../../../../Code/Handout Formatting}
          %   \newcommand{\stylesdir}{../../../../_styles/Yale_typeface}
          %
          %%%%%%%%%%%%%%%%%%%%%%
          
          \input{Handout Formatting/PDF Formatting Script.tex}
highlight-style: ../_styles/idle-fingers-light.theme
---

```{r}
#| include: false
#| echo: false

source("Code Highlighting Functions_Verbatim.R")

# Read all your bash snippets
bash_snippets <- list(
  "branch 'test-r' set up to track 'origin/test-r'.
Switched to a new branch 'test-r'"
)

# Generate all formatted code
formatted <- lapply(bash_snippets, format_bash)
#formatted <- lapply(bash_snippets, format_r)

# Write to a file you can copy-paste from
writeLines(unlist(formatted), "formatted_bash.tex")

```
This document applies code highlighting to Verbatim code environments, which are used to render formatted indentation in code blocks because this is not not natively supported in Quarto LaTeX documents. Below are the two steps to generate a formatted code block.

1. Paste the code to be highlighted into the bash_snippets list item. This will parse all provided code list items using the defined algorithm and generate a new TeX file named formatted_bash.tex. The resulting output may require manual adjustments, which can be made in the following two code block sections.


```{r}
#| include: false
#| echo: false

source("Code Highlighting Functions_Verbatim.R")

# Read all your bash snippets
bash_snippets <- list(
  "branch 'test-r' set up to track 'origin/test-r'.
Switched to a new branch 'test-r'"
)

# Generate all formatted code
formatted <- lapply(bash_snippets, format_bash)
#formatted <- lapply(bash_snippets, format_r)
#formatted <- lapply(bash_snippets, format_python)

# Write to a file you can copy-paste from
writeLines(unlist(formatted), "formatted_bash.tex")

```

2. Adjust the formatting by pasting the output below where Quarto will apply code highlighting, and in the following `\\begin{Verbatim}` section. The available commands are:

    - `\\kw` for Keywords
    - `\\func` for Functions
    - `\\str` for Strings
    - `\\com` for Comments
    - `\\op` for Operators
    - `\\atr` for Atributes


```{.bash}
git add .                      # Stage all changes
git commit -m "New plot"       # Commit staged changes
git push                       # Push to remote repository
```



\begin{list}{}{\setlength\leftmargin{2.5em}}
  \item\noindent\rule[0.3ex]{\dimexpr\textwidth - 2.5em\relax}{1pt}\vspace{-2.5ex}
  \item\small \textbf{Command-Line Interface} \normalsize \vspace{-2.5ex}
  \item\noindent\rule[0.3ex]{\dimexpr\textwidth - 2.5em\relax}{0.4pt}\vspace{-0.8ex}
  
  \begin{codebox}
  \begin{Verbatim}[breaklines = true, breaksymbol = \textcolor{idlelinnum}{\scriptsize$\hookrightarrow$}, commandchars = \\\{\}]
\func{git} add .                      \com{# Stage all changes}
\func{git} commit \op{-m} \str{"New plot"}       \com{# Commit staged changes}
\func{git} push                       \com{# Push to remote repository}
  \end{Verbatim}
  \end{codebox}

  \vspace{-4ex}
  \item\noindent\rule[0.3ex]{\dimexpr\textwidth - 2.5em\relax}{0.4pt}
\end{list}


