---
from: markdown+emoji
bibliography: references.bib
csl: ../../../Code/nature.csl
---

# `tidyverse` Worked-Through Example {.unnumbered}

<small>Updated: February 16<sup>th</sup>, 2026</small>

## Introduction

We will explore the `tidyverse` packages `tidyr`, `dplyr`, and `stingr` further with a worked-through example cleaning and visualizing COVID-19 daily death counts in the United States. The visualization will be prepared using another `tidyverse` package that was not discussed in this module. If you are interested in learning more about the `ggplot2` package, please check out our workshop on the topic: [Data Visualization with ggplot2](../../Data-Visualization-with-ggplot2/visualization-index.qmd).

## Set Up the Environment

First, we will load the necessary libraries and any special functions used in the script.

```{r}
#| message: FALSE
#| warning: FALSE

# NOTE: renv initialization might need to be run twice after the repo is
#       first copied.
#renv::init()
renv::restore()

suppressPackageStartupMessages({
  library("readr")      # For reading in the data
  library("tidyr")      # For tidying data 
  library("dplyr")      # For data manipulation 
  library("stringr")    # For string manipulation
  library("lubridate")  # For date manipulation
  library("ggplot2")    # For creating static visualizations
  library("scales")     # For formatting plots axis
})

# Function to select "Not In"
'%!in%' <- function(x,y)!('%in%'(x,y))
```

Now we will fetch the semi-cleaned COVID-19 deaths data set that has been prepared for this example.

::: {.callout-note collapse="true"}
## Data Source and Preparation Methods

The original, raw data comes from the Johns Hopkins University Coronavirus Resource Center (JHU CRC) Center for Systems Science and Engineering (CSSE) GitHub page: [CSSEGISandData](https://github.com/CSSEGISandData/COVID-19/tree/master). Details about these steps can be found in the `Intro-to-Programming-in-R/Code` directory ([link to code](https://github.com/ysph-dsde/Book-of-Workshops/tree/main/Workshops/Intro-to-Programming-in-R/Code)). The cleaned datasets are in the `Intro-to-Programming-in-R/Data` directory ([link to data](https://github.com/ysph-dsde/Book-of-Workshops/tree/main/Workshops/Intro-to-Programming-in-R/Data)).
:::

```{r}
# Save the GitHub raw URL associated with our data set
df_url  <- "https://raw.githubusercontent.com/ysph-dsde/Book-of-Workshops/refs/heads/main/Workshops/Intro-to-Programming-in-R/Data/Data%20for%20the%20Workshop_Aggregated%20by%20Month.csv"

# Read in the data set using the URL via the readr package
df_raw  <- read_csv(file = df_url, show_col_types = FALSE)
```

Before cleaning data, it's good practice to review several aspects of your raw data:

-   Review the variable Data Dictionary
-   Check the dataset dimensions (rows and columns)
-   Examine the first and/or last few rows
-   Check the variable type for each column

Most of these operations can be accomplished using dplyr's `glimpse()` function. Notice that only the first 7 columns are displayed here for readability; the full dataset contains 39 date columns.

```{r}
# Preview the raw data
glimpse(df_raw[, 1:7])
```

We can see that the dataset contains `r nrow(df_raw)` unique regions throughout the U.S. at the county- and state-level, with `r ncol(df_raw)` columns of information in total. The first 4 columns denote the geographic region in different formats, while the remaining 39 columns represent specific dates when COVID-19 deaths were added to the running tally. Importantly, all counts are reported as cumulative values (monotonically increasing) and have been summarized at monthly time intervals.

Regional identifier columns are classified as characters (`<chr>`), which is appropriate for text-based data. Death count columns are classified as double, meaning they are double-precision floating-point numbers—a numeric type that allows decimal values. Since deaths must be whole numbers (you can't have a fractional death), we need to convert these to integers, rounding down.

```{=html}
<div class="callout-discussion">
    <h2>
        <i class="fa-regular fa-comments"></i>
        Discussion Question
    </h2>
    <p>Is the data in a “tidy” format? Why or why not?</p>
</div> 
```

In the following sections, we will continue reviewing additional aspects of our dataset. Before we proceed, however, we need to tidy our data first.

## Clean Messy Data Using `tidyr`

Messy data might come in a multitude of forms, but the core data reshaping actions needed to achieve a tidy data format can be reduced to five operations:

- **Pivoting –** converts between long forms (observations added as new rows) and wide forms (observations added as new columns).
- **Rectangling –** turns deeply nested lists (as from JSON) into tidy tibbles, the tidyverse version of a dataframe object.
- **Nesting –** converts grouped data to a form where each group becomes a single row containing a nested data frame, and unnesting does the opposite.
- **Splitting/combining –** pulls a single character column into multiple columns or combines multiple columns into a single character column.
- **Managing `NAs` –** ensures that explicitly included `NAs` represent data that is actually missing or suppresses them so they are implicitly present instead.

`tidyr` offers multiple tools that achieve the same outcome, and these can be used interchangeably. However, sometimes one function will flow better in your script or read more intuitively.

As we saw in the previous section, observations are stored as columns, creating a wide dataset. To transform this into our target "tidy" format, columns need to represent only variables—not individual observations. We achieve this by converting date columns into rows, creating a long dataset. Fortunately, `tidyr` provides the perfect function for this transformation: `pivot_longer()`. We simply tell it which columns to pivot into rows.

```{r}
# Find the first and last dates in our array
str_c(c("Start date: ", "End date: "), colnames(df_raw)[c(5, ncol(df_raw))])
```

Using the start and end column names, we can pivot the date columns into a long format. We need to name two new columns for this operation: one to hold the pivoted column names (our dates), and another to hold their corresponding values. Note that the colon in "`r colnames(df_raw)[5]`:`r colnames(df_raw)[ncol(df_raw)]`" tells the function to include all consecutive columns between the start and end positions.

```{r}
df_long <- df_raw |> 
  # Step 1: Convert wide format date columns to long format
  pivot_longer(
    # Designate which columns need to be pivoted
    cols = "2020-01-01":"2023-03-01", 
    # Name the variable that will store newly pivoted column names
    names_to = "Date",
    # Name the variable that will store respective cell values
    values_to = "Deaths_Count_Cumulative"
  ) |> 
  # Step 2: Change table format to "data frame" class for convenience
  as.data.frame()

# Inspect the pivoted data
glimpse(df_long)
```

```{r}
#| echo: FALSE

# Report the dimensions of the pivoted data set
str_c(c("Number of rows: ", "Number of columns: "), c(nrow(df_long), ncol(df_long)) )
```

This looks much better! Each column now represents one variable, and each row represents one observation—matching a specific region with a specific date. Notice that we waited until now to change the death counts from doubles to integers. Why? Because with tidy data, this conversion is much simpler.

```{r}
# Convert double to integer
df_long$Deaths_Count_Cumulative <- as.integer(df_long$Deaths_Count_Cumulative)

# Use the Higher-Order function sapply() to get all variable classes
sapply(df_long, class) |> 
  # Convert from a vector to a data frame for easier reading
  as.data.frame() |> 
  # Change the column name for clarity
  `colnames<-`("Variable Class")
```

Technically, our dataset is now "tidy." But remember: tidy is just one step. We may still need additional cleaning and standardization. In the following sections, we'll continue preparing the data for visualization and analysis.

## Manipulate Data Using `dplyr`

While `tidyr` focuses on the mechanics of achieving the “tidy” format, this alone may not fully prepare your data. For example, you may need to transform or calculate new values from existing columns, combine datasets, or condense information unnecessarily spread across multiple columns. This process creates what's sometimes called the "minimally informative” form of a dataset—where all necessary values are included or derivable without redundancy or unnecessary information.

Functions available in the `dply` package is often used to create this “complete tidy” data frame.

Suppose we want to plot COVID-19 deaths across the entire U.S. First, we need to calculate total counts by summing deaths for each region (`Combined_Key`) on each date (`Date`). However, our dataset contains cumulative counts, which are difficult to interpret for visualization. We'll need to back-calculate monthly counts to show new deaths each month.

The `dplyr` package offers a suite of functions for data manipulation tasks like these. In the next four code blocks, we'll show how to generate counts for the entire U.S.

### Calculate Total U.S. Counts

Recall that the imported dataset includes values for U.S. counties, states, and territories. To calculate values for the U.S. as a whole, we need to focus only on the 50 states plus the District of Columbia. We can achieve this by filtering for `Combined_Key` entries that match those state names using the `filter()` function.

We're using `str_c()` from the `stringr` package here (string concatenate) to create a vector of state names for matching. We'll explore this function in detail in the next section.

**Step 1:** Filter subsets the data set by rows that match the condition.

```{r}
df_filtered <- df_long |> 
  filter(Combined_Key %in% str_c(c(datasets::state.name, "District of Columbia"), ", US"))
```

The full dataset has `r nrow(df_long)` rows. After filtering, we have `r nrow(df_filtered)` remaining—one for each U.S. state plus the District of Columbia across all months. Recall that observations were recorded over 39 months during the pandemic, so we expect `r nrow(df_filtered)`/39 = 51 unique regions.

Using this filtered dataset, we'll group rows by month using `group_by()`. The table below uses `tally()` to show how many rows are in each monthly group.

::: callout-tip
## Code Like a Pro

It's crucial to verify that your dataset is structured correctly and contains expected values. Since most datasets are too large for line-by-line inspection, use a different strategy: identify what features you expect in your data (like the 51 regions above) and verify them systematically.

This approach allows you to thoroughly validate that your dataset meets all expectations and requirements.
:::

**Step 2:** Group the table by unique entries in `Date`.

```{r}
df_grouped <- df_filtered |>
  group_by(Date)

# Display the number of entries associated with each group using tally()
df_grouped |> tally() |> 
  # Use an "anonymous function" to help pass the object from the
  # left side to the right side of the pipe
  (\(x) list("Tally" = head(x), "Unique n" = unique(x$n)))()
```

As expected, we have 51 row entries that are being grouped by each month recorded. Now we are ready to calculate the total U.S. counts. The function `summarize()` allows us to iterate a function over each grouping. If necessary, the function can be prompted to retain previously designated groups using `.groups = "keep"`.

**Step 3:** Calculate the cumulative deaths variable for the U.S. by summing over the grouped rows.

```{r}
df_US <- df_grouped |>
  summarize(Deaths_Count_Cumulative = sum(Deaths_Count_Cumulative), .groups = "keep")
```

Finally, we need to remove the groupings so we can work with the whole data set again.

**Step 4:** Remove grouping applied to the dataset object.

```{r}
df_US <- df_US |>
  ungroup()
```

After these four operations, we get the following data set:

```{r}
head(df_US)
```

Next, we'll merge this data back with the main dataframe using `bind_rows()`, a tidyverse function similar to base R's `do.call()` with notable differences. `bind_rows()` combines dataframes by matching column names, while `do.call()` appends data in the exact order given. Additionally, `bind_rows()` keeps all unique columns from all datasets being combined, filling missing columns with `NA`.

However, `df_US` is currently missing the `Combined_Key` and `Country_Region` columns. If we merge now, both will be filled with `NA`. Instead, we want them to contain "US." We can add these columns using `mutate()`, which creates new columns from functions of existing columns or static values.


```{r}
df_US <- df_US |>
  mutate(Combined_Key = "US", Country_Region = "US")

head(df_US)
```

Great, now that we have all the columns we need we are ready to merge it with `df_long`.

```{r}
df <- bind_rows(df_US, df_long)

head(df)
tail(df)
```

The `select()` function is used to keep, drop, or organize columns. In our case, we will use it to reorganize the columns. We want to arrange the columns such that we have the region first, followed by the month the record was made, and then the cumulative deaths recorded.

:::{.callout-important}
## Building Reproducible Workflows Through Good Documentation Practices

Remember that when cleaning data, you're preparing it not only for your use but potentially for others as well. Make your work well documented, sequentially organized, and easy to understand.
If you've ever kept a wet-lab notebook, think of how scientists document their work for future reference, collaborators, and auditors. Organizing rows and columns to reflect the logical progression of information improves readability and clarity of the final "cleaned tidy" format.
:::

```{r}
df <- df |>
  select(Combined_Key, County, Province_State, Country_Region, Date, Deaths_Count_Cumulative)

head(df)
```

### Back-Calculate Monthly Updates

Now we are ready to calculate the monthly death counts from the cumulative counts. Earlier, we mentioned that `mutate()` can be used to generate new columns as functions of existing columns. Here, we will generate a new column using `diff()` to back-calculate the values added each month.

```{r}
df <- df |> 
  # Group data by Combined_Key
  group_by(Combined_Key) |>
  # Ensure data is sorted by Date within each group
  arrange(Date) |>
  # Calculate daily monthly deaths counts using mutate
  mutate(Deaths_Count_Monthly =  Deaths_Count_Cumulative - lag(Deaths_Count_Cumulative, default = 0)) |>
  ungroup() |> 
  # Change table format to "data frame" for convenience
  as.data.frame()

head(df)
```

::: {.callout-tip collapse="true"}
## Code Like a Pro

It is important to ensure that your methods are accurate! For example, suppose you want to confirm the accuracy of directly calculating the cumulative counts for the U.S. over all 39 months. To do this, you can compare this direct approach (call this method "a") with an indirect method (call this method "b"). The indirect method first calculates the daily counts, sums these counts across all states, and then generates the cumulative counts.

```{r}
## --------------------
## Aggregate states directly from the cumulative counts
a <- df_filtered |>
  # Group data by Date
  group_by(Date) |>
  # Calculate the total cumulative deaths for each Date
  summarize(total_method_a = sum(Deaths_Count_Cumulative))

## --------------------
## First calculate the daily value, then recalculate the cumulative count
b <- df_filtered |>
  # Group data by Combined_Key
  group_by(Combined_Key) |>
  # Ensure data is sorted by Date within each group
  arrange(Date) |>
  # Calculate daily increments
  mutate(increment = Deaths_Count_Cumulative - lag(Deaths_Count_Cumulative, default = 0)) |>
  # Ungroup to remove group structure
  ungroup()

b <- b |>
  # Group data by Date
  group_by(Date) |>
  # Sum daily increments for each Date to get total increments
  summarize(total_increment = sum(increment)) |>
  # Calculate the cumulative sum of the total increments for each Date
  reframe(total_method_b = cumsum(total_increment)) |>
  ungroup()

## --------------------
## Compare the two methods to confirm they are the same. If they are, then the return
## value should be zero.
bind_cols(a, b) |>
  # Apply row-wise operations
  rowwise() |>
  # Calculate the absolute difference between the two cumulative totals
  summarize(diff = abs(total_method_a - total_method_b)) |>
  # Extract unique values of the differences
  (\(x) unique(x$diff))()
```

```{r}
#| echo: false

rm(a, b)
```
:::

## String Manipulation Using `stringr`

After completing data wrangling and tidying, it's good practice to review variable classifications and naming conventions before proceeding with calculations. For instance, if you have a variable for participant sex, ensure all entries are consistently labeled (e.g., "M" for Male, "F" for Female). Also confirm that zeros aren't used in place of `NA` for missing values, or vice versa.

`stringr` provides a comprehensive and consistently designed set of string manipulation functions based on `stringi`. It uses UTF-8, the standard encoding that includes most characters, and handles alternative encodings when needed.

While base R has similar functions, `stringr` offers key advantages:

- Consistent syntax across functions, seamless integration with tidyverse
- Regular updates to match operations in languages like Ruby or Python (base R updates more slowly)

:::{.callout-tip}
## Code Like a Pro

You can perform advanced operations using regular expressions, often referred to as “regex.” This workshop doesn't cover regex, but we encourage you to explore the additional resources provided earlier for an introduction to that topic.
:::

Let's start by examining the variable classifications again.

```{r}
sapply(df, class) |> 
  as.data.frame() |> 
  `colnames<-`("Variable Class")
```

The variables are mostly assigned to appropriate data classes. One exception: `Date` is a character but would be better as a date class for easier manipulation. We can convert it using the `lubridate` package from tidyverse. This package isn't covered in this introductory lecture, but you can review the documentation here: [lubridate](https://lubridate.tidyverse.org/).

The data is reported in`yyyy/mm/dd` format, so we use `ymd()` to convert the character to the date class.

```{r}
df$Date <- ymd(df$Date)
```

### Validate the Accuracy of Entries

For simplicity, we'll assume county-level entries are accurate and skip validation here, though in real analysis you'd want to verify this. We will, however, check that all U.S. states and territories appear in the state-level data by comparing unique `Combined_Key` values to `datasets::state.name` and the District of Columbia.

```{r}
# Find the non-states represented in the state-level data set
unique(df$Province_State)[unique(df$Province_State) %!in% c(datasets::state.name, "District of Columbia")]
```

`NA` values appear in the results. This makes sense because country-level aggregates don't have state values, so they show as `NA` at the state level. We can verify this to ensure all rows where `Province_State = NA` the `Combined_Key = "US"`.

```{r}
# Confirm assumption to check for possible source of irregularity
df[df$Province_State %in% NA, "Combined_Key"] |> unique()
```

Beyond the five U.S. territories, we also see entries for two cruise ships. Since these aren't relevant to our analysis, we'll exclude them using `str_detect()` or `str_which()`. Both functions identify rows matching a string pattern—`str_detect()` returns Boolean values (`TRUE`/`FALSE`) while `str_which()` returns row indices.

**Option #1:** Use the Boolean test to detect the presence of the "Princess" string and pull the matches. The added `%in% TRUE` expression ensures that `NA` outcomes are interpreted as `FALSE`.

```{r}
df_filtered <- df[!str_detect(df$Province_State, "Princess") %in% TRUE, ]
```

**Option #2:** Identify indices containing "Princess", then subset using the indices that don't.

```{r}
df_filtered <- df[-str_which(df$Province_State, "Princess"), ]
```

```{=html}
<div class="callout-discussion">
    <h2>
        <i class="fa-regular fa-comments"></i>
        Discussion Question
    </h2>
    <p>An alternative form of <code>str_which()</code> is to use its native negate setting: <code>str_which(df$Combined_Key, "Princess", negate = TRUE)</code>.

This works in many scenarios, but it's not appropriate here. Why?</p>
</div> 
```

This filtering removes the following number of rows, representing two unique regions with 39 dates each, as expected. 

```{r}
(nrow(df) - nrow(df_filtered))/2
```

### Generating New String Vectors

A common task is creating new string vectors from existing ones. Take `Combined_Key`, for example, which combines information from `County`, `Province_State`, and `Country_Region`. Suppose we only have county-level data and want to generate the corresponding state- and country-level `Combined_Key` values. We can approach this in two ways: using `str_c()` to combine separate columns, or using `str_split()` to extract components from existing strings.

For this example, we'll first subset to county-level data only. We can use `str_count()` to count how many times a pattern appears in each string. Since county-level data contains two commas, we'll filter using that pattern.

```{r}
df_state_level <- df_filtered[str_count(df_filtered$Combined_Key, ",") == 2, ]
```

**Option #1:** Generate a new column by combining the desired columns with `", "` as the separator.

```{r}
str_c(df_state_level$Province_State, df_state_level$Country_Region, sep = ", ") |> unique()
```

**Option #2:** Split the string only to the first observation of the deliminator.

:::{.callout-note}
By setting `n = 2`, we are restricting `str_split()` to return a maximum of 2 new vectors after the split operation. This means the function will split the string only at the first occurrence of the delimiter. Including the `simplify = TRUE` argument converts the output from a list (with each output as a separate list element) to a matrix, making it easier to handle the resulting data.
:::

```{r}
str_split(df_state_level$Combined_Key, ",", simplify = TRUE, n = 2) |>
  
  # Save the second column of information containing the state- and country-level
  # regional information. Notice the use of the underscore as a placeholder for 
  # the piped data.
  _[, 2] |>
  
  # Trim whitespace from both sides of the resulting strings
  str_trim(side = "both") |> unique()
```

### Standardizing Nomenclature

As a final step, we'll refine the naming for clarity and consistency. Suppose we want to specify that Virgin Islands entries refer only to the U.S. territory. We can do this using `str_replace()` to replace specific string patterns.

Going forward, we only require the `Combined_Key` variable, as the `Country_Region`, `Province_State`, and `County` variables are succinctly represented there. We can adjust the `"Virgin Islands"` entries for that column only.

```{r}
df_filtered[, "Combined_Key"] <- str_replace(df_filtered[, "Combined_Key"],  "Virgin Islands", "US Virgin Islands")
```

We can adjust multiple columns at once using the higher-order function `sapply()`. However, the base R pipe's \_ placeholder doesn't work with `sapply()`. To work around this, we wrap `sapply()` within the pipe and define our own placeholder for the incoming data. Here we use `x` within a temporary function. This temporary function is called an "anonymous function"—a concept we've used earlier in this example.

```{r}
# Identify the row indices where "Virgin Islands" is detected in "Province_State".
index = str_which(df_filtered[, c("Province_State")], "Virgin Islands")

df_filtered[index, c("Province_State", "Combined_Key")] |>
  # Define the anonymous function of this pipe-level and specify the information from
  # the left to be "x".
  (\(x) {
    # Apply the str_replace() function over "Province_State" and "Combined_Key".
    # Notice that the added "^ " is a regex statement that defines a hard
    # boundary on the string.
    sapply(x, function(y) str_replace(y,  "^Virgin Islands", "US Virgin Islands"))
  }) () |> 
  # Show that the changes have been completed.
  _[1:10, ]
```

Now we commit our changes to the primary dataframe label.

```{r}
df <- df_filtered
```

## Visualization: Daily COVID-19 Death Counts

To conclude the example, we will plot our tidy dataset and display the newly calculated country-wide counts.

```{r}

plot_deaths <- df |>
  # Select entries only reflecting country-level data
  filter(Combined_Key == "US") |>
  ggplot(aes(x = Date, y = Deaths_Count_Monthly)) +
    # Visualize as a line-plot
    geom_line(color = "#7634A6", linewidth = 1) +
    # Format the y-axis to show values in terms of thousands
    scale_y_continuous(labels = unit_format(unit = "K", scale = 1e-3)) +
    # Format the x-axis to show dates as Jan 2020 from 01/01/2020, spaced
    # every four months
    scale_x_date(date_breaks = "4 month", date_labels =  "%b %Y") +
    labs(x = "Date Recorded", y = "COVID-19 Deaths",
         title = "COVID-19 Deaths in the US - Grouped by Month") +
    theme_minimal()

plot_deaths
```

```{=html}
<div class="callout-discussion">
    <h2>
        <i class="fa-regular fa-comments"></i>
        Discussion Question
    </h2>
    <p>Work with your neighbors to answer the two questions in “<code>Questions.R</code>.”
    
    If you haven't downloaded the workshop code, see <a href="../programming-r-index.html#accessing-the-codespaces">Accessing the Codespace</a> for the code and setup instructions.</p>
</div> 
```

